---
title: "How Does a Computer Calculate sin(x)?"
author: "Yaroslav Vakula"
format: 
  dashboard:
    theme: cosmo
    orientation: columns
execute:
  echo: false
---

```{r setup, include=FALSE}
library(plotly)
library(dplyr)
library(tidyr)
library(knitr)

taylor_sin <- function(x, degree) {
  n_terms <- (degree + 1L) %/% 2L
  sapply(x, function(xi) {
    k <- seq_len(n_terms) - 1L
    sum((-1)^k * xi^(2*k + 1) / factorial(2*k + 1))
  })
}

x_vals  <- seq(-2 * pi, 2 * pi, length.out = 500)
degrees_ext <- seq(1, 13, by = 2)
colors_ext <- c("#E41A1C", "#FF7F00", "#4DAF4A", "#377EB8", "#984EA3", "#A65628", "#F781BF")
```

## Column {width=20%}

### Statistics {.sidebar}

This dashboard analyzes the approximation of $\sin(x)$ using 
**Maclaurin series**.

**Key Features:**

- Interactive visualization of polynomial convergence.

- Analysis across the interval $[-2\pi, 2\pi]$.

The Maclaurin series is defined as:
$$\sin(x) \approx \sum_{n=0}^{N} \frac{(-1)^n x^{2n+1}}{(2n+1)!}$$

**Technical Insight:**

For $n=13$, the approximation is virtually indistinguishable
from the actual function in the range $[-\pi, \pi]$. However, 
note how quickly the error grows outside this "safe" zone.

::: {.valuebox icon="calculator" color="primary"}
Target Function
`sin(x)`
:::

## Column {width=80%}

### Interactive Approximation {height=60%}
```{r}
#| title: "Convergence of Taylor Polynomials"

fig_approx <- plot_ly()

fig_approx <- add_trace(fig_approx,
  x    = x_vals,
  y    = sin(x_vals),
  type = "scatter",
  mode = "lines",
  name = "sin(x)",
  line = list(color = "black", width = 2.5, dash = "dash")
)

for (i in seq_along(degrees_ext)) {
  d <- degrees_ext[i]
  fig_approx <- add_trace(fig_approx,
    x    = x_vals,
    y    = taylor_sin(x_vals, d),
    type = "scatter",
    mode = "lines",
    name = paste0("n = ", d),
    line = list(color = colors_ext[i], width = 1.6)
  )
}

layout(fig_approx,
  xaxis = list(title = "x (radians)"),
  yaxis = list(title = "y", range = c(-3, 3)),
  hovermode = "x unified"
)
```
