---
title: "How Does a Computer Calculate sin(x)?"
author: "Yaroslav Vakula"
lang: en
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-title: "Content"
    number-sections: true
    code-summary: "Show code"
    highlight-style: github
    fig-align: "center"
    fig-width: 9
    fig-height: 6
    self-contained: true
    output-file: index.html
  pdf:
    toc: true
    toc-depth: 3
    toc-title: "Content"
    number-sections: true
    highlight-style: github
    fig-align: "center"
    fig-width: 9
    fig-height: 6
    output-file: index.pdf
execute:
  echo: false
crossref:
  eq-prefix: ""
---

## Introduction

Your calculator probably always knows the value of $\sin(x)$. 
But have you ever wondered how it calculates it exactly? A
computer, in the traditional sense, can't think geometrically
--- it doesn't understand angles, shapes, trigonometry, and so
on. All it can do is add, subtract, and multiply numbers, and 
it does this very well and quickly.


This is where polynomials come in. Since a polynomial uses
only these basic operations, any processor can easily
calculate it. Thus, if $\sin(x)$ is expressed as an arbitrary
polynomial, its calculation becomes simple.

Taylor series will help us with this. The Taylor series
represents any smooth function as an infinite sum of these
same polynomials. A particularly convenient special case is
the Maclaurin series, which is constructed around $x = 0$ and
works especially well when $x$ remains close to this point.

This paper examines how the Maclaurin series for $\sin(x)$ is
constructed, how accurate the approximations are, and how many
terms are actually needed to obtain a good answer.

## Theoretical Foundations

### Taylor Series

A Taylor series takes an arbitrary smooth function and
expresses it as an infinite sum, where each term is
constructed from the derivative of the function at some chosen
point $a$.

The general formula is shown in (@eq-taylor):

$$
f(x) = \sum_{n=0}^{\infty} \frac{f^{(n)}(a)}{n!}(x - a)^n
$$ {#eq-taylor}

Here $f^{(n)}(a)$ is the $n$-th derivative of $f$, evaluated at
point $a$ [@stewart2014calculus].

Each new term in such a sum slightly refines our
approximation. If we take enough such terms, we can get as 
close to the true value as we like. And since factorials grow 
very quickly, the terms quickly decrease for large $n$ --- that's 
why, in practice, only a few such terms are needed.

### Maclaurin Series

If we center the expansion at $a = 0$, we obtain a simpler and 
cleaner version of the series, called the Maclaurin series 
(@eq-maclaurin), which is actually used in most computers 
(microcontrollers, processors, etc.):

$$
f(x) = \sum_{n=0}^{\infty} \frac{f^{(n)}(0)}{n!} \, x^n
$$ {#eq-maclaurin}

Since each derivative is evaluated at $0$, the calculation is 
significantly simplified. The main drawback of this approach 
is that the accuracy of the approximation decreases as $|x|$ 
increases --- that is, the further the desired value is from 
zero, the more terms you need to maintain the desired accuracy.

### Expansion of $\sin(x)$

$\sin(x)$ has a pleasant structural property that makes its 
series particularly pure. Its derivatives pass through only 
four forms --- $\sin$, $\cos$, $-\sin$, $-\cos$ --- and at 
$x = 0$, each sine-based derivative vanishes ($\sin(0) = 0$), 
while the cosine-based derivatives equal $\pm 1$. This leaves 
only odd powers of $x$.

The result is the Maclaurin expansion [@stewart2014calculus]:

$$
\sin(x) = x - \frac{x^3}{3!} + \frac{x^5}{5!} - \frac{x^7}{7!} 
+ \cdots
$$ {#eq-sine}

The signs alternate and the factorials in the denominators
increase rapidly --- meaning that each subsequent term is much 
smaller than the previous one, especially near zero.

Of course, you can never calculate an infinite number of 
terms, and in practice, computing infinitely many terms is 
impossible. At some point, the calculation stops, and only 
a remainder remains. This remainder indicates how large the 
error of given approximation is. That is, how much the 
approximation deviates from the true result, which can be 
obtained analytically rather than numerically.

## Approximation Error

If $f(x)$ is the real function and $T_n(x)$ is a Taylor 
polynomial of degree $n$, then the error is simply the 
difference between them:

$$
R_n(x) = f(x) - T_n(x)
$$ {#eq-error}

To bound this error without adding an infinite number of 
terms, we use the Lagrange form of the remainder [@wiki_taylor]. 
For the Maclaurin series (where $a = 0$) it looks like this:

$$
R_n(x) = \frac{f^{(n+1)}(\xi)}{(n+1)!} x^{n+1}
$$ {#eq-lagrange}

where $\xi$ is some real number between $0$ and $x$.

This is particularly useful because all derivatives of 
$\sin(x)$ are either $\pm\sin(x)$ or $\pm\cos(x)$, both of 
which have absolute values bounded by $1$. So the inequality 
simplifies nicely [@wiki_taylor]:

$$
|f^{(n+1)}(\xi)| \leq M \implies |f^{(n+1)}(\xi)| \leq 1
$$ {#eq-estimate}

This means that the absolute error satisfies:

$$
|R_n(x)| \le M \left| \frac{x^{n+1}}{(n+1)!} \right| 
\implies |R_n(x)| \le \left| \frac{x^{n+1}}{(n+1)!} \right|
$$ {#eq-sine-bound}

Two things immediately follow:

* When $x$ is close to zero, $x^{n+1}$ is small --- hence the 
error is small.

* As $n$ increases, the factorial $(n+1)!$ increases sharply, 
which reduces the error very quickly.

## Visualization

### $\sin(x)$ and Taylor Polynomials

@fig-approx shows $\sin(x)$ as a dashed black line alongside 
five Taylor polynomial approximations of degrees $1$ through 
$13$. Degree 1 is simply the tangent line $y = x$. Each higher 
degree adds another term from (@eq-sine), and the approximation 
improves significantly over a larger range.

```{r}
#| label: setup
#| echo: false

library(ggplot2)
library(dplyr)
library(plotly)

taylor_sin <- function(x, degree) {
  n_terms <- (degree + 1L) %/% 2L
  sapply(x, function(xi) {
    k <- seq_len(n_terms) - 1L
    sum((-1)^k * xi^(2*k + 1) / factorial(2*k + 1))
  })
}

x_vals  <- seq(-2 * pi, 2 * pi, length.out = 500)
degrees <- c(1, 3, 5, 7, 9)
pal     <- c(
  "#E41A1C", 
  "#FF7F00", 
  "#4DAF4A", 
  "#377EB8", 
  "#984EA3"
  )
```

```{r}
#| label: fig-approx
#| fig-cap: "Graph: sin(x) and Taylor polynomial approximations for degrees 1 to 13"

degrees_ext <- seq(1, 13, by = 2)

colors_ext <- c(
  "#E41A1C", "#FF7F00", "#4DAF4A",
  "#377EB8", "#984EA3", "#A65628", "#F781BF"
)

if (knitr::is_html_output()) { # render for html (interactive)
  fig_approx <- plot_ly()

  fig_approx <- add_trace(fig_approx,
    x    = x_vals,
    y    = sin(x_vals),
    type = "scatter",
    mode = "lines",
    name = "sin(x)",
    line = list(color = "black", width = 2.5, dash = "dash")
  )

  for (i in seq_along(degrees_ext)) {
    d <- degrees_ext[i]
    fig_approx <- add_trace(fig_approx,
      x    = x_vals,
      y    = taylor_sin(x_vals, d),
      type = "scatter",
      mode = "lines",
      name = paste0("n = ", d),
      line = list(color = colors_ext[i], width = 1.6)
    )
  }

  layout(fig_approx,
    xaxis = list(
      title    = "x",
      tickvals = list(-2*pi, -pi, 0, pi, 2*pi),
      ticktext = list("-2\u03c0", "-\u03c0", "0", "\u03c0", "2\u03c0")
    ),
    yaxis  = list(title = "y", range = c(-3, 3)),
    legend = list(x = 1.01, y = 0.99, xanchor = "left"),
    margin = list(r = 130)
  )

} else { # render for pdf (static)
  library(ggplot2)
  library(tidyr)

  df <- data.frame(x = x_vals, "sin(x)" = sin(x_vals), check.names = FALSE)
  for(i in seq_along(degrees_ext)) {
    df[[paste0("n = ", degrees_ext[i])]] <- taylor_sin(x_vals, degrees_ext[i])
  }
  
  df_long <- pivot_longer(df, -x, names_to = "curve", values_to = "y")
  df_long$curve <- factor(df_long$curve, levels = colnames(df)[-1])

  ggplot(df_long, aes(x, y, color = curve, linetype = curve, linewidth = curve)) +
    geom_hline(yintercept = 0, color = "black", linewidth = 0.4) +
    geom_vline(xintercept = 0, color = "black", linewidth = 0.4) +
    geom_line() +
    scale_color_manual(values = c("black", colors_ext)) +
    scale_linetype_manual(values = c("dashed", rep("solid", length(degrees_ext)))) +
    scale_linewidth_manual(values = c(1.2, rep(0.7, length(degrees_ext)))) +
    scale_x_continuous(
      breaks = c(-2*pi, -pi, 0, pi, 2*pi), 
      labels = expression(-2*pi, -pi, 0, pi, 2*pi)
    ) +
    coord_cartesian(ylim = c(-3, 3)) +
    labs(x = "x", y = "y") +
    theme_minimal() +
    theme(legend.title = element_blank())
}
```

Close to $x = 0$, degree $13$ is almost indistinguishable from 
a true sinusoidal curve --- it only starts to deviate at the 
edges of the graph. Polynomials of lower degrees lose their 
coherence much earlier.

## Convergence Analysis

The Maclaurin series for $\sin(x)$ converges everywhere --- 
for any value you assign to $x$. But how quickly it converges 
is another matter, and depends strongly on how far $x$ is from 
zero.

Near zero, successive terms decrease rapidly, and only a few 
are needed. Further away from zero, the initial terms are 
large and tend to partially cancel out, so the partial sums 
take much longer to settle close to the true value.

@tbl-convergence shows the minimum degree of the polynomial 
required to achieve three different accuracy goals at six 
control points.

```{r}
#| label: tbl-convergence
#| tbl-cap: "Minimum polynomial degree required to approximate sin(x) within the given accuracy"
#| echo: false

find_min_degree <- function(x_val, tol) {
  n <- 1
  while (TRUE) {
    approx <- taylor_sin(x_val, n)
    
    exact_err <- abs(sin(x_val) - approx)
    
    if (exact_err < tol) {
      return(paste0("$n = ", n, "$"))
    }
    
    n <- n + 2
  }
}

x_vals_target <- c(0.5, 1.0, 2.0, pi, 2*pi, 3*pi)
x_labels      <- c("0.5", "1.0", "2.0", "$\\pi$", "2$\\pi$", "3$\\pi$")

col_e3 <- sapply(x_vals_target, find_min_degree, tol = 1e-3)
col_e6 <- sapply(x_vals_target, find_min_degree, tol = 1e-6)
col_e12 <- sapply(x_vals_target, find_min_degree, tol = 1e-12)

conv_df_dynamic <- data.frame(
  x      = x_labels,
  acc_e3 = col_e3,
  acc_e6 = col_e6,
  acc_e12 = col_e12
)

knitr::kable(
  conv_df_dynamic,
  col.names = c(
    "$x$", 
    "Accuracy $10^{-3}$", 
    "Accuracy $10^{-6}$", 
    "Accuracy $10^{-12}$"
    ),
  align     = c("c", "c", "c", "c"),
  escape    = FALSE
)
```

To understand why convergence slows down with distance from 
zero, consider how successive terms relate to one another. The 
general term of the series is:

$$
a_k = \frac{x^{2k+1}}{(2k+1)!}
$$ {#eq-general-term}

Taking the ratio of consecutive terms:

$$
\frac{a_{k+1}}{a_k} = \frac{x^{2k+3}}{(2k+3)!} \cdot 
\frac{(2k+1)!}{x^{2k+1}} = \frac{x^2}{(2k+3)(2k+2)}
$$ {#eq-term-ratio}

Substituting $n = 2k+1$ for the degree of the current term, 
the denominator becomes $(n+2)(n+1)$, which for large $n$ is 
well approximated by $n^2$. This gives:

$$
\frac{a_{k+1}}{a_k} \approx \frac{x^2}{n^2}
$$ {#eq-ratio-approx}

Each new term is therefore smaller than the previous one by a 
factor of roughly $\frac{x^2}{n^2}$. When $x$ is small, this 
ratio is tiny and the series collapses quickly. When $x$ is 
large, the ratio stays close to $1$ for many terms before 
shrinking --- which is precisely why more terms are needed far 
from zero.

In real processors, the Maclaurin series is never applied 
directly to large inputs. Instead, a range reduction is first 
applied: by using the identity $\sin(x + 2\pi) = \sin(x)$, any 
input is mapped into a small interval near zero. From there, 
just a few terms are enough to reach full floating-point 
precision --- keeping the computation both fast and accurate.

## Conclusion

So, how does a computer calculate $\sin(x)$? The answer is the 
Maclaurin series (@eq-sine) --- an infinite sum where each 
term is simply an odd power of $x$ divided by a factorial. 
Nothing more exotic than that.

Near $x = 0$, the series converges quickly and a small number 
of terms is all you need. Further out, more terms are required 
--- which is exactly why real implementations lean on range 
reduction, pulling every input back toward zero before the 
series is ever evaluated.

The accuracy gain isn't linear, either. Each additional term
cuts the error by roughly a factor of (@eq-ratio-approx). Near 
zero that factor is small, so convergence is rapid; further 
away, it's larger, so things slow down. @tbl-convergence 
captures this behavior concretely.

Taylor and Maclaurin series show up across physics, 
engineering, signal processing, machine learning --- really, 
anywhere that complex functions need to be computed 
efficiently. What they reveal is one of the more quietly 
powerful ideas in numerical mathematics: that even the most 
intricate functions can be pinned down using nothing but 
addition and multiplication. That's a genuinely elegant result.

## References

::: {#refs}
:::
