---
title: "How Does a Computer Calculate sin(x)?"
author: "Yaroslav Vakula"
format:
  revealjs:
    theme: simple
    transition: slide
    slide-number: true
    fig-align: "center"
    output-file: slides.html

execute:
  echo: false
---

## The Problem

Your calculator always returns a value for $\sin(x)$.

. . .

But how does it actually compute it?

. . .

- A computer **cannot** reason geometrically
- It only knows how to **add, subtract, and multiply**
- It needs an arithmetic recipe --- not a geometric one

## The Key Idea: Polynomials

A polynomial uses only addition, subtraction, and 
multiplication:

$$p(x) = a_0 + a_1 x + a_2 x^2 + \cdots + a_n x^n$$

. . .

**Any processor can evaluate a polynomial efficiently.**

. . .

So if $\sin(x)$ can be written as a polynomial --- the problem 
is solved.

## Taylor Series

Any smooth function can be written as an infinite sum of 
polynomial terms [@stewart2014calculus]:

$$f(x) = \sum_{n=0}^{\infty} \frac{f^{(n)}(a)}{n!}(x - a)^n$$

. . .

- $f^{(n)}(a)$ --- the $n$-th derivative of $f$ at point $a$
- Each new term **refines** the approximation
- Factorials grow fast $\rightarrow$ terms shrink fast $\rightarrow$ few terms needed in practice

## Maclaurin Series

Setting $a = 0$ gives the **Maclaurin series** --- the form used 
in most computing environments:

$$f(x) = \sum_{n=0}^{\infty} \frac{f^{(n)}(0)}{n!} \, x^n$$

. . .

**Advantages:**

- All derivatives evaluated at $0$ $\rightarrow$ simpler coefficients
- Very accurate when $|x|$ is small

. . .

**Drawback:** accuracy decreases as $|x|$ grows

## Expansion of $\sin(x)$

The derivatives of $\sin(x)$ cycle through four forms:
$$\sin \to \cos \to -\sin \to -\cos \to \cdots$$

. . .

At $x = 0$: sine-based terms vanish $(\sin(0) = 0)$, cosine-based equal $\pm 1$.

Only **odd powers** survive:

$$\sin(x) = x - \frac{x^3}{3!} + \frac{x^5}{5!} - \frac{x^7}{7!} + \cdots$$

## Approximation Error

In practice, we stop after $n$ terms. The error is:

$$R_n(x) = f(x) - T_n(x)$$

. . .

The **Lagrange remainder** [@wiki_taylor] gives an upper bound:

$$R_n(x) = \frac{f^{(n+1)}(\xi)}{(n+1)!} x^{n+1}, \quad \xi \in (0,\, x)$$

. . .

$$|R_n(x)| \le \left| \frac{x^{n+1}}{(n+1)!} \right|$$

## What the Error Bound Tells Us

$$|R_n(x)| \le \left| \frac{x^{n+1}}{(n+1)!} \right|$$

. . .

**Near zero** ($|x|$ small): $x^{n+1}$ is tiny $\rightarrow$ error is tiny

. . .

**Adding more terms** (larger $n$): $(n+1)!$ grows explosively 
$\rightarrow$ error shrinks fast

. . .

Both effects work in our favour near $x = 0$.

## Visualization {.smaller}

```{r}
#| label: setup

library(ggplot2)
library(dplyr)
library(plotly)

taylor_sin <- function(x, degree) {
  n_terms <- (degree + 1L) %/% 2L
  sapply(x, function(xi) {
    k <- seq_len(n_terms) - 1L
    sum((-1)^k * xi^(2*k + 1) / factorial(2*k + 1))
  })
}

x_vals <- seq(-2 * pi, 2 * pi, length.out = 500)
```

```{r}
#| label: fig-approx
#| fig-cap: "sin(x) and Taylor polynomial approximations for degrees 1 to 13"

degrees_ext <- seq(1, 13, by = 2)
colors_ext  <- c("#E41A1C","#FF7F00","#4DAF4A","#377EB8","#984EA3","#A65628","#F781BF")

fig_approx <- plot_ly()

fig_approx <- add_trace(fig_approx,
  x = x_vals, y = sin(x_vals),
  type = "scatter", mode = "lines", name = "sin(x)",
  line = list(color = "black", width = 2.5, dash = "dash")
)

for (i in seq_along(degrees_ext)) {
  d <- degrees_ext[i]
  fig_approx <- add_trace(fig_approx,
    x = x_vals, y = taylor_sin(x_vals, d),
    type = "scatter", mode = "lines",
    name = paste0("n = ", d),
    line = list(color = colors_ext[i], width = 1.6)
  )
}

layout(fig_approx,
  xaxis  = list(title = "x",
                tickvals = list(-2*pi, -pi, 0, pi, 2*pi),
                ticktext = list("-2π", "-π", "0", "π", "2π")),
  yaxis  = list(title = "y", range = c(-3, 3)),
  legend = list(x = 1.01, y = 0.99, xanchor = "left"),
  margin = list(r = 130)
)
```

## How Many Terms Are Needed? {.smaller}

```{r}
#| label: tbl-convergence
#| tbl-cap: "Minimum polynomial degree to approximate sin(x) within a given accuracy"

find_min_degree <- function(x_val, tol) {
  n <- 1
  while (TRUE) {
    if (abs(sin(x_val) - taylor_sin(x_val, n)) < tol)
      return(paste0("$n = ", n, "$"))
    n <- n + 2
  }
}

x_vals_target <- c(0.5, 1.0, 2.0, pi, 2*pi, 3*pi)
x_labels      <- c("0.5", "1.0", "2.0", "$\\pi$", "2$\\pi$", "3$\\pi$")

knitr::kable(
  data.frame(
    x       = x_labels,
    acc_e3  = sapply(x_vals_target, find_min_degree, tol = 1e-3),
    acc_e6  = sapply(x_vals_target, find_min_degree, tol = 1e-6),
    acc_e12 = sapply(x_vals_target, find_min_degree, tol = 1e-12)
  ),
  col.names = c("$x$", "$10^{-3}$", "$10^{-6}$", "$10^{-12}$"),
  align = c("c","c","c","c"),
  escape = FALSE
)
```

The required degree **grows with** $|x|$.

## Why Convergence Slows Down

The ratio of consecutive terms explains the pattern:

$$\frac{a_{k+1}}{a_k} = \frac{x^2}{(2k+3)(2k+2)}
 \approx \frac{x^2}{n^2}$$

. . .

- **Small** $x$: ratio $\ll 1$ $\rightarrow$ series collapses in a few 
terms
- **Large** $x$: ratio $\approx 1$ for many terms $\rightarrow$ slow 
convergence

. . .

This is why real processors use **range reduction** first:

$$\sin(x + 2\pi) = \sin(x)$$

## Conclusion

**How does a computer calculate $\sin(x)$?**

. . .

$$\sin(x) = x - \frac{x^3}{3!} + \frac{x^5}{5!} - \frac{x^7}{7!} + \cdots$$

. . .

- Only odd powers of $x$, divided by factorials --- nothing 
exotic
- Highly accurate near $x = 0$ with very few terms
- Range reduction extends this to all inputs

. . .

**The big idea:** even the most complex functions can be 
pinned down using nothing but addition and multiplication.

## References

::: {#refs}
:::
